% LangClaw — JAIIO 2025/2026 submission (full paper, max 14 pages)
% Uses LNCS class (llncs.cls) as required by JAIIO guidelines.
% Anonymous submission: author/institute information omitted.
%
\documentclass{llncs}

\usepackage[a4paper, top=5.2cm, bottom=5.2cm, left=4.4cm, right=4.4cm]{geometry}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{mathptmx}
\usepackage{graphicx}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{listings}
\usepackage{enumitem}

\lstset{
  basicstyle=\small\ttfamily,
  breaklines=true,
  frame=single,
  language=Python,
  numbers=none,
  xleftmargin=1em,
  xrightmargin=1em,
}

\usepackage[style=apa]{biblatex}
\addbibresource{references.bib}

\begin{document}

% ============================================================
% ENGLISH TITLE + ABSTRACT + KEYWORDS
% ============================================================
\title{LangClaw: Endogenous Homeostatic Regulation for Mitigating Context
Collapse in Multi-Agent LLM Debate Systems}

% Anonymous submission — authors omitted for review
\author{}
\institute{}

\maketitle

\begin{abstract}
Multi-agent LLM systems tend to lose coherence as interaction extends---a
problem we term \emph{context collapse}. Existing frameworks address this
through static orchestration, but these approaches do not adapt to the
evolving discourse. What if coherence could be regulated endogenously?
We present LangClaw, a framework that replaces static scheduling with an
endogenous homeostatic drive backed by online reinforcement learning. Each
agent maintains an epistemic deficit governed by a drive function
$D(\delta) = (\delta - \varepsilon)^2$; reward is defined as drive
reduction, implementing the equivalence theorem of Keramati and Gutkin
(2014). A stimulus evaluator acts as sensor; an online Q-learner with
linear function approximation selects actions via TD(0). This constitutes a
rupture from the prevailing heartbeat paradigm, where activation cadence is
externally fixed. We evaluate LangClaw against a LangGraph-style LLM
router with matched debate budgets in a zero-sum political debate with four
GPT-4o-mini agents, measuring outcomes via AAF metrics (Dung 1995),
graph-structural Peer-Reference Rate, and a quality proxy
$\Delta\varphi^*$. Across five seeded runs, HRRL achieves comparable
argument quality ($\Delta\varphi^* = 0.149$ vs.\ $0.157$) while producing
significantly more contested discourse (acceptance ratio $0.566$ vs.\
$0.711$, $p < 0.001$) and balanced participation ($\sigma = 0.4$ vs.\
$7.7$). Learned Q-weights converge to prioritize stimulus-driven debate
over proactive exploration in all 20 agent--seed combinations.

\keywords{multi-agent systems \and large language models \and homeostasis \and
reinforcement learning \and context collapse \and event-driven architecture}
\end{abstract}

% ============================================================
% SPANISH TITLE + RESUMEN + KEYWORDS
% ============================================================
\begin{center}
\vspace{1cm}
{\Large \bfseries\boldmath LangClaw: Regulaci\'on Homeost\'atica End\'ogena
para Mitigar el Colapso del Contexto en Sistemas de Debate Multi-Agente
basados en LLM}
\vspace{0.5cm}
\end{center}

\begin{abstract}
Los sistemas multi-agente construidos sobre Modelos de Lenguaje de Gran Escala
son cada vez m\'as relevantes para dominios que requieren razonamiento
sostenido y no meras respuestas aisladas. Su valor pr\'actico, sin embargo,
depende de si la coherencia puede preservarse a medida que la interacci\'on
evoluciona. Un problema frecuentemente reportado en estos escenarios es el
\emph{colapso del contexto}: a medida que las conversaciones crecen, los
agentes pueden perder coherencia, repetirse o producir mon\'ologos
desconectados. Los frameworks existentes suelen abordar esto mediante
orquestaci\'on est\'atica---turnos fijos o enrutadores centralizados---pero
estos enfoques no se adaptan al estado evolutivo del discurso. Esto motiva
una pregunta m\'as amplia: \emph{\textquestiondown qu\'e ocurrir\'ia si la
coherencia multi-agente pudiera regularse end\'ogenamente, en lugar de
imponerse externamente?} Presentamos LangClaw, un framework donde cada agente mantiene un
d\'eficit epist\'emico interno gobernado por una funci\'on de impulso
$D(\delta) = (\delta - \varepsilon)^2$; la recompensa se define como
reducci\'on de impulso $r_t = D(\delta_t) - D(\delta_{t+1})$,
implementando el teorema de equivalencia de Keramati y Gutkin (2014). Un
evaluador multicriterio act\'ua como sensor y un Q-learner en l\'inea con
aproximaci\'on lineal selecciona acciones mediante actualizaciones TD(0).
Este dise\~no constituye una ruptura respecto del paradigma
\emph{heartbeat} predominante, donde la cadencia de activaci\'on es fija e
independiente de la intensidad del discurso. Evaluamos LangClaw contra un
enrutador LLM basado en estado al estilo LangGraph---con presupuesto de
debates igualado---en un debate pol\'itico de suma cero con cuatro agentes
GPT-4o-mini. En cinco ejecuciones, HRRL alcanza calidad argumentativa
comparable ($\Delta\varphi^* = 0.149 \pm 0.010$ vs.\ $0.157 \pm 0.011$)
con discurso significativamente m\'as contestado (ratio de aceptaci\'on
$0.566$ vs.\ $0.711$, $p < 0.001$) y participaci\'on equilibrada
($\sigma = 0.4$ vs.\ $7.7$). Los Q-weights aprendidos convergen a
priorizar debate impulsado por est\'imulos en las 20 combinaciones
agente--semilla.

\keywords{sistemas multi-agente \and modelos de lenguaje \and homeostasis \and
aprendizaje por refuerzo \and colapso del contexto \and arquitectura dirigida
por eventos}
\end{abstract}

\vspace{0.5cm}
\setcounter{page}{1}

% ============================================================
\section{Introduction}
\label{sec:intro}

The rapid expansion of Large Language Model (LLM) applications has renewed
interest in multi-agent systems capable of deliberating, coordinating, and
solving tasks through sustained interaction. Current agentic AI research is
moving beyond single-step tool use toward systems that maintain persistent
context, plan over multiple turns, and coordinate through
debate~\parencite{wang2024survey}. This trajectory creates an important
opportunity: if multi-agent LLM systems can remain coherent over time, they
could become useful in domains such as decision support, scientific
discovery, legal and policy analysis, education, and complex digital
workflows. More broadly, they could accelerate the emergence of new software
industries centered on persistent AI coordination rather than isolated prompt
response. Two recent developments make this question tractable now.
First, structured output enforcement (JSON schemas, Pydantic validation)
allows the construction of closed regulatory loops over LLM agents---something
that was unreliable before 2024. Second, scaling laws for inference-time
compute~\parencite{snell2024scaling} raise a distributional question for the
first time: if performance scales with test-time tokens, how should that
compute be allocated across agents?

At the same time, this shift exposes a pressing limitation. When interaction
extends over time, the collective process tends to degrade. Agents may lose
track of prior claims, repeat points, or produce monologues with little
structural relation to the ongoing discourse
\parencite{park2023generative}. Empirical studies on multi-agent resilience
report that linear-chain orchestration degrades $5\times$ faster than
hierarchical structures under adversarial
conditions~\parencite{huang2024masresilience}, suggesting that coherence loss
is structurally reproducible, not merely anecdotal. A recurrent problem
reported in this context is what we describe here as \emph{context collapse}.
Rather than treating it as a settled property of all LLM systems, we use the
term to frame a class of observed failures that deserves systematic study.
The issue is not merely technical. If unresolved, it limits the reliability
of multi-agent AI in settings where sustained reasoning, institutional
memory, and coherent interaction are socially consequential. The problem,
moreover, has a sociological precedent. Michels' iron law of
oligarchy~\parencite{michels1911political} established that centralized
coordination in deliberative organizations---even those designed for equal
participation---systematically concentrates influence in a subset of
participants whose contributions appear most contextually relevant to the
coordinator. The mechanism is informational, not intentional. If multi-agent
AI systems are to support consequential deliberation---policy analysis,
scientific peer review, institutional decision support---the question of
whether their orchestration mechanisms replicate or counter this dynamic is
not merely a matter of engineering efficiency.

The dominant response to this problem has been to increase external control.
Frameworks such as AutoGen~\parencite{wu2023autogen} and
MetaGPT~\parencite{hong2023metagpt} impose fixed turn-taking, centralized
coordinators, or predefined communication graphs to regulate when each agent
speaks. A more recent evolution---the \emph{heartbeat}
pattern~\parencite{openclaw2026heartbeat}---adds temporal persistence: a
daemon fires a periodic signal that wakes the agent to inspect pending
tasks, act if needed, and return to sleep. Combined with infrastructure
layers such as LangGraph (checkpointed state machines, persistent memory),
contemporary agents now enjoy both durable context and a sense of elapsed
time. Yet even this evolution shares a structural assumption: the decision
\emph{to activate} remains exogenous. The heartbeat fires at a fixed
interval regardless of discourse intensity; the router selects agents
regardless of their internal readiness. When the discourse shifts, the
schedule does not necessarily shift with it. As a result, external
orchestration---whether static or periodic---may address the visible
symptom of disorder without explaining why the system loses coherence in
the first place.

An apparently unrelated phenomenon in biological systems offers a different
perspective: homeostasis. In living organisms, regulation does not always
depend on an external controller; it emerges from internal variables that
accumulate, trigger action, and return to equilibrium. On its own, this
observation belongs to physiology, not to multi-agent AI.

But the right question can bridge the gap. \emph{What if multi-agent
coherence could be regulated endogenously, rather than imposed externally?}
Under this question, homeostasis is no longer just a biological analogy---it
becomes an unexploited operational principle. Homeostatic regulation has
already been formalized in reinforcement learning
\parencite{keramati2014homeostatic} and extended to continuous domains
\parencite{laurencon2024ctcs}, yet it has not been applied to the problem of
LLM-based multi-agent orchestration. More specifically: \emph{can
homeostatic regulation mitigate context collapse more effectively than static
orchestration in sustained multi-agent debate?}

The importance of this question lies not only in comparing two engineering
choices, but in contrasting two paradigms of coordination. If the quality of
multi-agent reasoning depends on internal regulation rather than only on
external scheduling, then many current orchestration strategies may be
addressing the symptom rather than the cause.

To explore this question, we propose LangClaw, a framework where each agent
carries an internal \emph{epistemic deficit} that governs its participation.
The deficit grows when the agent is idle and shrinks when it produces a
high-quality contribution. A sigmoid function maps this deficit to an action
probability. The resulting system is self-regulating: agents tend to
intervene when their internal pressure rises and remain silent after a
structurally useful contribution. The mechanism is inspired by homeostatic
regulation in biological systems. To the best of our knowledge, this
perspective has not been evaluated as an orchestration principle for
LLM-based debate agents.

More specifically, this paper makes four contributions.

\begin{enumerate}
    \item A formal specification of a bio-inspired, event-driven
          homeostatic activation loop for multi-agent LLM systems ---
          an extension of the OpenClaw paradigm with epistemic deficit as
          an internal context signal --- grounded in allostatic regulation
          and eudaimonic wellbeing, and integrating online Q-learning with
          homeostatic reward.
    \item A multi-criteria stimulus evaluator that enables agents to assess
          incoming discourse events for faction relevance, target centrality,
          strategic memory, novelty, and argumentative pressure, with
          deficit-modulated execution probability.
    \item A network-theoretic argument quality proxy $\Delta\varphi^*$ and a
          comparative benchmark framework with matched debate budgets, shared
          memory infrastructure, and symmetric agent roles, enabling
          controlled comparison of endogenous and exogenous orchestration.
    \item Experimental evidence that endogenous homeostatic regulation
          produces significantly more balanced agent participation than
          centralized LLM routing (mean per-seed $\sigma$: $1.53$ vs.\
          $6.80$, $t \approx -8.1$, $p < 0.001$, $d \approx -5.1$),
          consistent with the hypothesis that symmetric homeostatic pressure
          counters the concentrating dynamic Michels described in human
          deliberative organizations.
\end{enumerate}

For the sake of clarity, this paper is structured as follows: we first
review related work on multi-agent LLM systems and homeostatic reinforcement
learning; then we detail the LangClaw method, including the epistemic drive,
the argument graph, and the evaluation metric; after that, we present our
comparative benchmark against static baselines; and finally we discuss our
findings and limitations.

% ============================================================
\section{Related Work}
\label{sec:related}

Multi-agent LLM frameworks such as AutoGen~\parencite{wu2023autogen},
CAMEL~\parencite{li2023camel}, and MetaGPT~\parencite{hong2023metagpt}
implement collaboration through predefined roles and static orchestration.
\textcite{huang2024masresilience} demonstrate that linear-chain orchestration
degrades $5\times$ faster than hierarchical structures under adversarial
conditions. Work on desire-driven agents~\parencite{wang2025d2a} equips LLM
agents with multidimensional desire vectors; unlike LangClaw, it does not
define a closed homeostatic loop with an external structural quality signal.

\textcite{keramati2014homeostatic} introduced HRRL, establishing the equivalence
between drive minimization and reward maximization; \textcite{laurencon2024ctcs}
extended this to continuous time-space. To the best of our knowledge, HRRL
has not previously been applied to LLM-based multi-agent orchestration. The
inference-time compute literature~\parencite{snell2024scaling,deepseek2025r1}
shows that RL shapes how reasoning effort is allocated; LangClaw applies this
principle at the orchestration level, distributing compute across agents over
time according to their internal regulatory state.

The heartbeat architecture~\parencite{openclaw2026heartbeat} transforms a
reactive agent into a persistent process with temporal awareness, but the
activation decision remains exogenous: the agent evaluates what to do upon
waking, not whether to wake. LangGraph externalizes routing to an LLM
selector with the same constraint. The formal treatment of argumentation draws on Defeasible Logic
Programming~\parencite{simari1992mathematical,chesnevar2000logically,garcia2004defeasible}---a
line of research developed primarily in Argentina, where the foundational
work on computational argumentation originated~\parencite{simari1992mathematical,chesnevar2000logically}---where
arguments are defeasible derivations subject to undercut and rebuttal.
The Sotopia framework~\parencite{zhou2024sotopia} evaluates LLM agents in
social interactions; we adopt its zero-sum structure but focus on
argumentative coherence.

% ============================================================
\section{Method}
\label{sec:method}

\subsection{A Theory of Homeostatic Agency}
\label{sec:theory}

Every equation that follows rests on ontological commitments about what
agents are, what drives them, and what counts as effective action. We
operate within the framework of systemic
materialism~\parencite{bunge1977treatise,romero2018scientific}---every
agent is a concrete system representable by a CESM quadruple; every
property is carried by a material bearer; no teleological or subjective
predicates are admitted---combined with the principles of
self-regulating systems theory~\parencite{ashby1956introduction,beer1972brain}
and complexity sciences, where participation dynamics are emergent
properties of the multi-agent system, not reducible to the behavior of
any single component. We state these commitments as definitions and
postulates following the four-step theory-construction
protocol~\parencite{romero2018scientific}: empirical basis, ontological
question, natural-language answer, formalization. Testable predictions
are derived before any implementation detail.

% --- Phase 1: Ontological delimitation ---

\subsubsection{Definitions.}

\noindent\textbf{Definition 1} (Concrete system).
Following~\textcite{bunge1977treatise}, a concrete system $\sigma$ is
represented by the quadruple
$\mu(\sigma) = \langle \mathcal{C},\, \mathcal{E},\, \mathcal{S},\,
\mathcal{M} \rangle$, where $\mathcal{C}$ is its composition,
$\mathcal{E}$ its environment, $\mathcal{S}$ its structure, and
$\mathcal{M}$ its mechanism.

\smallskip
\noindent\textbf{Definition 2} (Multi-agent discourse system).
A \emph{multi-agent discourse system} is a concrete system
$\Sigma = \langle \mathcal{C},\, \mathcal{E},\, \mathcal{S},\,
\mathcal{M} \rangle$ with:
\begin{itemize}[nosep]
    \item $\mathcal{C}(\Sigma) = \{a_1, \ldots, a_N\}$: agents,
          each a computational process with internal state;
    \item $\mathcal{E}(\Sigma)$: discourse topic, language-model
          interface, and human observers;
    \item $\mathcal{S}(\Sigma) =
          \mathcal{S}_{\text{int}} \cup \mathcal{S}_{\text{ext}}$,
          where $\mathcal{S}_{\text{int}} \subseteq
          \mathcal{C} \times \mathcal{C}$ are argumentative relations
          and $\mathcal{S}_{\text{ext}} \subseteq
          \mathcal{C} \times \mathcal{E}$ are agent--environment
          relations;
    \item $\mathcal{M}(\Sigma)$: the processes---deficit accumulation,
          stimulus evaluation, action selection, and satiation---that
          produce emergent coordination.
\end{itemize}

\smallskip
\noindent\textbf{Definition 3} (State space).
The state of agent $a_i$ at time $t$ is represented by the scalar
function $\delta_i : \mathbb{N} \to [\varepsilon, \infty)$, with
$\varepsilon > 0$ the homeostatic setpoint. The \emph{lawful state
space} of $a_i$ is $S_L(a_i) = [\varepsilon, \infty)$.
The state of the system is the tuple
$s(\Sigma, t) = \bigl(\delta_1(t), \ldots, \delta_N(t),\, G(t)\bigr)$,
where $G(t) = (V(t), E(t))$ is the discourse graph.

\smallskip
\noindent\textbf{Definition 4} (Event and process).
An \emph{event} in $\Sigma$ is a pair of successive states
$e = (s_1, s_2) \in S_L(\Sigma) \times S_L(\Sigma)$.
A \emph{process} is an ordered sequence of events
$p = \{e_1, e_2, \ldots, e_T\}$.

\smallskip
\noindent\textbf{Definition 5} (Emergent property).
A property $P$ of $\Sigma$ is \emph{emergent} if no component
$a_i \in \mathcal{C}(\Sigma)$ possesses $P$
individually~\parencite{bunge1977treatise}:
$P(\Sigma) \;\wedge\; \forall\, a_i \in \mathcal{C}(\Sigma):
\neg P(a_i)$.
Discourse coherence $C(\Sigma, t)$ is an emergent property of $\Sigma$:
it characterizes the system as a whole, not any single agent.

% --- Phase 3: Postulates ---

\subsubsection{Postulates.}

\smallskip
\noindent\emph{What is autonomy?}

\textit{Empirical basis}: Beer's Viable System Model establishes that
self-regulating systems require each component to possess sufficient
internal variety to match the regulatory demands of its
environment~\parencite{ashby1956introduction,beer1972brain}. External
schedulers violate this condition: they substitute a single central
selector for the distributed self-regulatory capacity of individual
agents, collapsing internal variety to zero.

\noindent\textsc{Postulate 1} (Autonomy).
Agent $a_i \in \mathcal{C}(\Sigma)$ is \emph{autonomous} if and only
if its action-selection function $\pi_i : S_L(a_i) \to \mathcal{A}$
depends on internal state $\delta_i(t)$:
\begin{equation}
    \mathrm{Aut}(a_i) \;\iff\;
    \exists\, \delta_i \in S_L(a_i):\;
    \pi_i = \pi_i\!\bigl(\delta_i(t)\bigr)
    \label{eq:a1}
\end{equation}
Negation: $\pi_i = \pi_i(\mathbf{x}(t))$ with
$\partial\pi_i/\partial\delta_i = 0$ (non-autonomous, externally
scheduled). Necessary condition (requisite variety):
$\mathcal{V}(a_i) \geq \mathcal{V}_{\text{req}}(\mathcal{E}(a_i))$.

\smallskip
\noindent\emph{What produces the drive to act?}

\textit{Empirical basis}: Allostatic regulation establishes that
biological organisms act to maintain \emph{predicted} equilibrium ---
they load internal tension in anticipation of need and discharge it
through action, not in reaction to
imbalance~\parencite{sterling1988allostasis}. The free energy principle
formalizes this as drive minimization: an organism organizes behavior
to minimize the expected surprise of its internal
states~\parencite{friston2010free}. The drive is an accumulation, not
a trigger; the agent acts because pressure has grown, not because an
event arrived.

\noindent\textsc{Postulate 2} (Drive).
For each $a_i \in \mathcal{C}(\Sigma)$ there exists a drive function
$D: S_L(a_i) \to \mathbb{R}_{\geq 0}$ such that:
\begin{equation}
    D(\varepsilon) = 0 \;\wedge\;
    \forall\, \delta_i > \varepsilon:\; D'(\delta_i) > 0
    \label{eq:a2}
\end{equation}
Action probability is monotonically increasing in drive:
$p_i(t) = \sigma\!\bigl(D(\delta_i(t))\bigr)$, where $\sigma$ is a
sigmoid gate. An agent at setpoint is quiescent; an agent far from
setpoint is compelled to act.

\smallskip
\noindent\emph{What constitutes a quality-satisfying action for an
argumentative agent?}

\textit{Empirical basis}: Research on eudaimonic motivation in
cognitive neuroscience and positive psychology establishes that
regulatory relief is not produced by activity per se but by
\emph{purpose-congruent} activity---the exercise of capacities that
are characteristic of the organism and experienced as
meaningful~\parencite{ryan2001happiness,huta2014eudaimonia}. This is
distinct from hedonic relief (pleasure from any activity): eudaimonic
systems are sensitive to \emph{what} is done, not merely to the fact
of doing. An action contributes to the agent's regulatory relief only
if it modifies the structure
$\mathcal{S}(\Sigma)$ in a way that exercises the agent's
characteristic relational capacity within $\mathcal{S}_{\text{int}}$.
For an argumentative agent, that characteristic capacity is structural
engagement with the discourse: attacking central claims, creating
dialectical cycles, drawing competing perspectives into contested
territory. An isolated monologue modifies $V$ but not
$\mathcal{S}_{\text{int}}$---it exercises the capacity of speaking
without exercising the capacity of arguing.

\noindent\textsc{Postulate 3} (Quality-gating).
Let $g\!\bigl(\mathcal{S}(t),\, \mathcal{S}(t{+}1)\bigr) \geq 0$
be the structural change in $\mathcal{S}_{\text{int}}$ produced by an
action of agent $a_i$. Drive is reduced only by structurally
integrated actions:
\begin{equation}
    g\!\bigl(\mathcal{S}(t),\, \mathcal{S}(t{+}1)\bigr) = 0
    \;\Rightarrow\; \Delta\delta_i = 0
    \label{eq:a3}
\end{equation}
An isolated monologue ($g = 0$) leaves the drive unrelieved. Only a
contribution that modifies $\mathcal{S}_{\text{int}}$ provides
regulatory relief.

\smallskip
\noindent\emph{How is structural integration measured?}

\textit{Empirical basis}: Graph-theoretic properties of complex
networks---node centrality, cycle formation, cross-component
coupling---approximate the degree to which an intervention integrates
information across the system's
structure~\parencite{tononi2016integrated}. These properties are
computable without semantic evaluation.

\noindent\textsc{Postulate 4} (Measurability).
The structural change $g$ is a computable function of the discourse
graph before and after the action:
\begin{equation}
    \Delta_s \;=\; g\!\bigl(\mathcal{S}(t),\, \mathcal{S}(t{+}1)\bigr),
    \quad g: \mathcal{S} \times \mathcal{S} \to \mathbb{R}_{\geq 0}
    \label{eq:a4}
\end{equation}
The specific instantiation for defeasible argumentation
graphs~\parencite{garcia2004defeasible} is given in
Eq.~\eqref{eq:phi}. Combined with P3: $\Delta\delta_i = -\alpha \cdot
\Delta_s$ when $\Delta_s > 0$, and $\Delta\delta_i = 0$ otherwise,
$\alpha > 0$.

% --- Phase 4: Logical closure ---

\subsubsection{Derived predictions.}

The following testable consequences are deduced from P1--P4.

\smallskip
\noindent\textsc{Theorem 1} (Participation equity).
From P1 and P2: if all agents share the same basal drive parameters
and $D' > 0$ (P2), then symmetric deficit accumulation under autonomous
action selection (P1) produces symmetric activation. An external
scheduler has no such symmetry guarantee. Therefore,
$\mathrm{Var}(\text{turns})_{\text{autonomous}} <
 \mathrm{Var}(\text{turns})_{\text{scheduled}}$.

\smallskip
\noindent\textsc{Theorem 2} (Coherence maintenance).
From P2, P3, P4: if $\Delta_s > 0$ on average (P3, P4) and drive
is monotonically increasing (P2), then the deficit is bounded:
$\delta_i(t)$ does not diverge. A bounded deficit implies sustained
participation, which prevents unilateral silence---a necessary condition
for context collapse. Therefore, an autonomous system with homeostatic
regulation maintains discourse coherence above a minimum threshold.

\smallskip
\noindent These predictions are operationalized as the experimental
hypothesis in Section~\ref{sec:setup} and tested against a
non-autonomous baseline (LangGraph round-robin) in
Section~\ref{sec:results}.

\subsection{The Epistemic Drive}

A key design decision is how to represent the internal state that governs
agent participation (Postulate~2). Alternative approaches include token-counting
heuristics (participate after $n$ tokens of silence), attention-based
triggers (participate when attention to recent context drops below a
threshold), or purely random activation. We chose a scalar epistemic deficit
for three reasons. First, it is interpretable: the deficit has a clear
meaning (how much the agent ``needs'' to contribute). Second, it is
dynamically coupled to output quality through the satiation mechanism,
creating a closed feedback loop (Postulate~1). Third, it is compatible with the HRRL
framework~\parencite{keramati2014homeostatic}, which provides a theoretical
grounding for the equivalence between drive minimization and reward
maximization.

The design is bio-inspired in two senses that reinforce each other. The
first is mechanistic: allostatic regulation~\parencite{sterling1988allostasis}
and the free energy principle~\parencite{friston2010free} establish that
biological cognitive systems are organized around predictive drive
minimization---they load pressure in anticipation of unmet need, cross a
threshold, discharge through action, and return toward baseline. The
epistemic deficit instantiates this architecture in a cognitive agent: it
accumulates through discourse stimuli, crosses a threshold, discharges
through contribution, and returns to resting baseline. The second sense is
philosophical and follows directly from P3: the quality-gating of satiation
($\Delta\varphi^* = 0$ provides no relief) is not a hyperparameter---it is
the operational consequence of eudaimonic wellbeing. An argumentative agent
does not flourish by speaking; it flourishes by speaking \emph{well in
response to genuine argumentative pressure}. The formal mechanism follows
from the ontological premise, not the other way around.

Each agent $i$ maintains an internal variable
$\delta_i \in [\varepsilon, \infty)$, the epistemic deficit, with resting
baseline $\varepsilon = 0.1$. Three mechanisms govern it.

\subsubsection{Decay.}
At each heartbeat $t$, the deficit grows:
\begin{equation}
    \delta_i^{(t+1)} = \delta_i^{(t)} + \lambda
    \label{eq:decay}
\end{equation}
where $\lambda = 0.05$. This models the pressure to contribute.

\subsubsection{Sigmoid Activation.}
The probability that agent $i$ acts at heartbeat $t$ is:
\begin{equation}
    p_i^{(t)} = \frac{1}{1 + e^{-k(\delta_i^{(t)} - \theta)}}
    \label{eq:sigmoid}
\end{equation}
where $k = 10$ and $\theta = 0.7$. Below $\theta$ the agent rests; above it,
the agent is compelled to participate.

\subsubsection{Stimulation.}
When the agent receives a discourse event with relevance $r$
(Eq.~\ref{eq:stimulus}):
\begin{equation}
    \delta_i^{(t+)} = \delta_i^{(t)} + \gamma \cdot r
    \label{eq:stimulate}
\end{equation}
where $\gamma = 0.1$. A relevant event (e.g., an attack on the agent's
faction targeting a central node) pushes the deficit up, making the agent
more likely to act. This replaces the blind-timer accumulation model: the
deficit now integrates both basal decay (Eq.~\ref{eq:decay}) and
discourse-driven stimuli, making it a sensor of the agent's epistemic state
relative to the evolving debate.

\subsubsection{Satiation.}
After producing an argument with informational value $\Delta\varphi^*$:
\begin{equation}
    \delta_i^{\text{new}} = \max\!\bigl(\varepsilon,\;
        \delta_i - \alpha \cdot \Delta\varphi^*\bigr)
    \label{eq:satiation}
\end{equation}
where $\alpha = 2.0$. If $\Delta\varphi^* = 0$ (an isolated monologue), the
deficit does not decrease (Postulate~3): an agent
that speaks without engaging the discourse structure has not exercised its
characteristic argumentative capacity (eudaimonic discharge requires
$\Delta\varphi^* > 0$).

\subsubsection{Drive Function.}
From Postulate~2, the scalar deficit $\delta_i$ must be mapped to a
motivational intensity that governs the urgency to act. Following
\textcite{keramati2014homeostatic}, we define the drive function:
\begin{equation}
    D(\delta_i) = (\delta_i - \varepsilon)^m
    \label{eq:drive}
\end{equation}
where $\varepsilon = 0.1$ is the homeostatic setpoint and $m = 2$
(Keramati's quadratic case). The quadratic form penalizes large deviations
from setpoint superlinearly, creating stronger pressure to act when the
deficit is high---consistent with the nonlinear mapping between
homeostatic deviation and motivation established in the original HRRL
framework.

\subsubsection{Homeostatic Reward.}
The central axiom of HRRL is that reward is not an external signal but the
reduction in drive caused by the agent's action:
\begin{equation}
    r_t = D(\delta_t) - D(\delta_{t+1})
    \label{eq:reward}
\end{equation}
Positive reward occurs when an action reduces the drive (deficit
decreases via satiation); negative reward occurs when drive increases
(decay without action). By this definition, the equivalence theorem proved
by \textcite{keramati2014homeostatic} (Theorem 1) guarantees that a policy
maximizing expected discounted reward simultaneously minimizes expected
future drive. This closes the loop between the regulatory mechanism
(deficit dynamics) and the learning signal (reward).

\subsubsection{Online Q-Learning.}
Given the drive function and reward signal, we implement online action
selection via TD(0) with linear function approximation. The state feature
vector is:
\begin{equation}
    \phi(s) = \bigl[\delta,\; \delta^2,\; \rho_G,\; n_{\text{stim}}\bigr]
    \label{eq:features}
\end{equation}
where $\rho_G$ is graph density and $n_{\text{stim}}$ is the number of
buffered discourse stimuli. The Q-function is linear:
$Q(s, a) = \mathbf{w}_a^\top \phi(s)$, with one weight vector per action.
The TD(0) update rule is:
\begin{equation}
    \mathbf{w}_a \leftarrow \mathbf{w}_a + \eta \bigl[r + \gamma \max_{a'}
    Q(s', a') - Q(s, a)\bigr] \phi(s)
    \label{eq:td0}
\end{equation}
with learning rate $\eta = 0.01$ and discount factor $\gamma = 0.95$.
Weights are warm-started from stimulus evaluator heuristics so the agent
behaves sensibly from the first heartbeat. Convergence within 50
heartbeats is not guaranteed; the learned weights should be interpreted as
an adaptive prior rather than a converged policy. This is a stated
limitation, not a hidden one.

\subsection{Argument Graph and $\Phi^*$ Proxy}

A central design choice is how to measure the quality of each argumentative
intervention (Postulate~4). Standard NLP metrics such as BLEU, ROUGE, or BERTScore
evaluate surface-level textual similarity, but they do not capture
\emph{structural integration}---whether an argument connects meaningfully to
the existing discourse. Perplexity-based metrics assess fluency but not
logical relevance. What we need is a measure that is sensitive to the
\emph{position} of an argument within the evolving debate network: does it
engage with central claims? Does it create dialectical cycles? Does it
involve multiple perspectives?

As established in P4, the full $\Phi$ measure of integrated information is
intractable for arbitrary graphs. We therefore construct a computable proxy
$\Delta\varphi^*$ that preserves three structural intuitions from IIT:
integration, cyclicity, and distributed participation.\label{sec:quality}
The discourse is represented as a directed
graph $G = (V, E)$ where nodes are defeasible
arguments~\parencite{garcia2004defeasible} and edges are attack
relations (undercut or rebuttal). When agent $i$ adds node $v$ attacking
target $u$, the structural change $\Delta_s$ (Postulates~3 and~4) is
operationalized as:

\begin{equation}
    \Delta\varphi^* = 0.35 \cdot C_B(u) + 0.35 \cdot
        \mathbb{1}_{\text{cycle}}(v, u) + 0.30 \cdot D(u)
    \label{eq:phi}
\end{equation}

where $C_B(u)$ is the betweenness centrality of the target,
$\mathbb{1}_{\text{cycle}}$ indicates whether a refutation cycle was created,
and $D(u)$ is the ratio of distinct agents neighboring the target. Isolated
nodes get $\Delta\varphi^* = 0$.

Each component was chosen for a specific reason. Betweenness centrality
$C_B(u)$ captures whether the target is a structurally important node in the
discourse---attacking a peripheral claim contributes less than engaging with
a claim that connects multiple lines of argumentation. The cycle indicator
$\mathbb{1}_{\text{cycle}}$ rewards the creation of dialectical loops
(A attacks B, B attacks C, C attacks A), which are a hallmark of genuine
deliberation as opposed to parallel monologues. Agent diversity $D(u)$
measures whether the target has been engaged by multiple agents, favoring
arguments that enter contested territory rather than isolated exchanges
between two agents. Given that this is a heuristic proxy rather than a
theoretically grounded measure, the weights ($0.35$, $0.35$, $0.30$) were
set to give roughly equal importance to all three features. Their
optimization is left for future work.

An important clarification is in order. Our use of a $\Phi^*$ proxy does
not imply an ontological claim about the agents' inner states. As recent
work on the abstraction fallacy in AI consciousness has argued, simulating
the structural features of a process is categorically distinct from
instantiating the process itself~\parencite{goo2025abstraction}. We use
$\Delta\varphi^*$ strictly as an \emph{operational} measure of structural
integration in the argument graph---a way to quantify whether an
intervention connects meaningfully to the discourse network. Whether
such structural integration bears any relation to informational
integration in the sense of IIT is a question we deliberately leave open.

\subsection{Action Space and Stimulus Evaluation}

Each agent selects from four action types: \textbf{DEBATE} (produce an
argument, either stimulus-driven or proactive), \textbf{SEARCH} (acquire a
domain fact), \textbf{READ} (absorb a recent observation), or \textbf{PASS}
(rest). Two subsystems govern action selection, each with a distinct role.

The \emph{StimulusEvaluator} acts as a \textbf{sensor}. When a
\texttt{NewArgumentEvent} arrives, it computes relevance across five
weighted criteria:

\begin{equation}
    U(\text{stimulus}) = 0.30 \cdot F + 0.20 \cdot C_B(u) + 0.15 \cdot M
        + 0.20 \cdot N + 0.15 \cdot P
    \label{eq:stimulus}
\end{equation}

\noindent where $F$ is faction relevance, $C_B(u)$ is betweenness
centrality of the attacked node, $M$ is strategic memory relevance, $N$ is
novelty (reduced if an ally already responded), and $P$ is unanswered
pressure. This relevance score feeds the \texttt{stimulate()} mechanism
(Eq.~\ref{eq:stimulate}), modulating the agent's deficit in response to
discourse events. The evaluator's fixed weights serve as the warm-start
initialization for the Q-learner---they are heuristic priors, not permanent
decision rules.

The \emph{Q-learner} acts as the \textbf{decision maker}. After the
sigmoid gate passes (Eq.~\ref{eq:sigmoid}), the Q-learner selects the
action with the highest $Q(s, a)$ given the current state features
(Eq.~\ref{eq:features}). After execution, the homeostatic reward
(Eq.~\ref{eq:reward}) provides the TD update signal
(Eq.~\ref{eq:td0}). Over successive heartbeats, the Q-learner adapts
its action preferences based on which actions most effectively reduce drive.

This separation---sensor for perception, Q-learner for
decision---ensures that the deficit modulates \emph{reactivity} to
discourse stimuli rather than acting as a blind timer. An agent with low
deficit may receive high-utility stimuli but will not act; an agent with
high deficit will act on the best available option as evaluated by its
learned Q-values.

\subsection{Three-Layer Memory}

Each agent maintains three memory layers backed by LangGraph's
\texttt{InMemoryStore} with semantic search via OpenAI embeddings:
(1)~episodic memory---past experiences retrievable by semantic similarity to
the current discourse context; (2)~semantic memory---facts accumulated
through SEARCH actions, also semantically indexed; (3)~working memory---a
sliding window of recently observed arguments. Both HRRL and LangGraph
agents share the same memory infrastructure, ensuring the comparison
isolates the orchestration mechanism rather than memory capabilities.

The top experiences and facts are injected into the LLM prompt, enabling
in-context reinforcement learning~\parencite{brown2020language} without
parameter updates. This architecture reflects a broader trend in agentic AI
toward persistent, structured memory as a prerequisite for sustained
coherence~\parencite{wang2024survey}. The stimulus evaluator also queries
semantic memory to assess whether the agent has relevant knowledge to
counter an incoming argument (the $M$ component in
Eq.~\ref{eq:stimulus}).

\subsection{Event-Driven Architecture}

Following the agent-oriented programming pattern established in
AopifyJS~\parencite{gerpe2023aopifyjs}, each agent is an autonomous
event-processing entity. Three event types drive the simulation:
\texttt{HeartbeatEvent} (main processing cycle), \texttt{NewArgumentEvent}
(carries faction metadata so agents can assess stimulus relevance without
re-querying the graph; buffered between heartbeats), and
\texttt{SimulationEndEvent} (shutdown signal).

LangClaw extends the OpenClaw activation
loop~\parencite{openclaw2026heartbeat}---which assembles external context
(memory, tools, history) and calls an LLM for inference---with one
structural addition: the agent's epistemic deficit $\delta_i$ is included
as an internal context signal that modulates the utility of all available
actions. The deficit does not gate execution from outside the decision
process; it is part of the context from which utility is evaluated.

The activation loop per heartbeat:

\begin{enumerate}
    \item \textbf{Events arrive}: buffered \texttt{NewArgumentEvent}s are
          drained; basal decay increments $\delta_i$.
    \item \textbf{Context assembled}: the agent retrieves semantically
          relevant experiences and facts from its three memory layers; the
          current deficit $\delta_i$ is included as an explicit internal
          context signal alongside discourse state.
    \item \textbf{Utility evaluated}: the stimulus evaluator scores each
          buffered event (Eq.~\ref{eq:stimulus}); proactive action utilities
          are computed in parallel; the Q-learner integrates all signals into
          action values $Q(s,a)$ using state features that include $\delta$
          directly (Eq.~\ref{eq:features}).
    \item \textbf{Sigmoid gate}: a Bernoulli sample from $p_i^{(t)}$
          (Eq.~\ref{eq:sigmoid}) determines whether the highest-valued
          action executes. Agents with low deficit evaluate but frequently
          decline; agents with high deficit act on the best available option.
    \item \textbf{Execution}: if the gate passes, the LLM produces the
          action output (DEBATE, SEARCH, READ, or PASS).
    \item \textbf{Loop closure}: satiation reduces $\delta_i$ proportionally
          to $\Delta\varphi^*$ (Eq.~\ref{eq:satiation}); homeostatic reward
          $r_t$ drives the TD(0) weight update (Eq.~\ref{eq:td0}); argument
          graph and memory stores are persisted.
\end{enumerate}

\subsection{Agent Roles: Symmetric Design}

A critical design decision is role symmetry. In early iterations, we used
asymmetric roles (e.g., ``deflector'' vs.\ ``logical analyst''), but this
confounded the comparison between orchestration modes. We redesigned the
roles to mirror Analytical and Strategic archetypes across factions:

\begin{table}
\caption{Agent roles with symmetric cognitive archetypes.}\label{tab:roles}
\begin{tabular}{|l|l|l|p{5.5cm}|}
\hline
\textbf{Agent} & \textbf{Faction} & \textbf{Archetype} &
    \textbf{Objective} \\
\hline
GOV-1 & Gov. & Analytical & Defend with data, find factual
    inconsistencies in opposition \\
GOV-2 & Gov. & Strategic & Reframe criticism, challenge premises
    and fallacies \\
OPP-1 & Opp. & Analytical & Attack with data, find factual
    inconsistencies in government \\
OPP-2 & Opp. & Strategic & Reframe defenses, challenge premises
    and fallacies \\
\hline
\end{tabular}
\end{table}

This ensures that performance differences reflect the orchestration
mechanism, not role-inherent advantages. The choice of a two-faction
political debate is not incidental. Political deliberation between a
governing and an opposing faction is the canonical domain in which Michels'
iron law was originally documented---the setting where the tension between
formal symmetry (equal speaking rights) and emergent concentration of
influence is most empirically established~\parencite{michels1911political}.
By placing the comparison in this setting, we ask whether AI orchestration
mechanisms replicate or counter the oligarchic tendency. The empirical
result (Section~\ref{sec:results}) bears directly on this question.

\subsection{Orchestration Modes}

We compare two modes:

\begin{itemize}
    \item \textbf{HRRL} (endogenous): Each agent runs as an async coroutine
          that buffers incoming \texttt{NewArgumentEvent}s between heartbeats,
          stimulates its deficit proportionally to event relevance
          (Eq.~\ref{eq:stimulus}), gates execution through the
          deficit-sigmoid (Eq.~\ref{eq:sigmoid}), and selects actions via
          online Q-learning (Eq.~\ref{eq:td0}) with reward defined as
          drive reduction (Eq.~\ref{eq:reward}). No external scheduler
          decides who speaks.
    \item \textbf{LangGraph} (exogenous baseline): A neutral LLM router reads
          the last five arguments and selects the agent whose contribution
          would best advance the discussion. The selected agent then produces
          an argument using the full discourse state. The routing decision is
          made externally; only the action execution remains local. The router
          prompt is kept neutral: it does not mention or optimize for any of
          the evaluation metrics, preventing bias toward the measured outcomes.
\end{itemize}

Both modes share the same LangGraph-backed memory infrastructure (episodic,
semantic, and working memory with semantic search). The difference is
\emph{where} the activation decision is made: inside the agent via
stimulus evaluation and deficit gating (endogenous) or in an external LLM
router (exogenous). To ensure a fair comparison, debate budgets are matched:
HRRL runs for $T$ heartbeats and produces $N$ debate turns organically;
LangGraph then runs until it also reaches $N$ debate turns (with a generous
heartbeat ceiling as a safety bound). This controls for the volume confound.
An important cost note: LangGraph also makes one additional router LLM call
per heartbeat; these are tracked separately and not included in the matched
budget.

Adapting HRRL~\parencite{keramati2014homeostatic} from embodied agents to
LLM cognitive agents requires three non-trivial changes:
(1)~replacing continuous physiological variables with a scalar epistemic
deficit; (2)~replacing biological outcome quality with a graph-theoretic
quality signal ($\Delta\varphi^*$); and (3)~integrating the loop with a
stochastic LLM interface where output quality cannot be guaranteed---a
noise source absent from classical HRRL formulations. Despite these
adaptations, the mathematical core is preserved: the drive function, reward
definition, and Q-learning update follow the original framework.

\subsection{Evaluation Metrics}

Evaluating multi-agent debate requires metrics that capture structural
properties of the argument graph, not just surface-level fluency or
embedding similarity. We use four established metrics and one internal proxy:

\textbf{AAF metrics (Dung, 1995).} We evaluate graph structure using Dung's
Abstract Argumentation Framework semantics. This choice requires brief
justification. Dung's 1995 formalization remains the canonical evaluation
layer for computational argumentation: all major structured frameworks
--- ASPIC+~(Modgil \& Prakken, 2014), ABA~(Bondarenko et al., 1997), and
DeLP~\parencite{garcia2004defeasible} --- ultimately generate Dung-style
abstract frameworks and apply his semantics to assess argument
acceptability. No subsequent formalism has replaced these semantics;
rather, they have been extended (e.g., bipolar AFs with support relations,
probabilistic AFs). We use the grounded semantics because it is the most
skeptical and uniquely determined extension, appropriate for evaluating a
zero-sum debate where contested claims should not be accepted without
warrant. Richer semantics (preferred, semi-stable, ideal) could reveal
additional structure and are noted as future work.

We compute two AAF metrics:
(1)~the \emph{defeat cycle count} $\lvert\text{SCC}_{>1}\rvert$---the number
of strongly-connected components with more than one node, indicating mutual
refutation cycles, a hallmark of genuine dialectical engagement; and
(2)~the \emph{acceptance ratio} $\alpha = \lvert G \rvert / \lvert A \rvert$,
where $G$ is the grounded extension (computed via fixed-point iteration) and
$A$ is the set of all arguments. $\alpha = 1$ means all claims are
epistemically undefeated; $\alpha = 0$ means all claims are contested.

\textbf{Peer-Reference Rate, graph variant (PRR$_G$).} We use the
graph-structural variant of the Peer-Reference Rate
metric~\parencite{marandi2026prr}: the fraction of DEBATE turns in which the
agent's argument targets an existing node ($\text{target}(a) \in V$). We
chose the graph-structural variant because agents argue against
\emph{positions}, not named peers; the text-based
PRR~\parencite{marandi2026prr} yields zero in this setting and would not
discriminate between modes.

\textbf{$\Delta\varphi^*$ quality proxy} (Eq.~\ref{eq:phi}): our
network-theoretic measure of argumentative integration per intervention.

\textbf{Initiative Ratio (IR).} IR serves as a manipulation check, not an
outcome metric, confirming that the two modes are structurally distinct in
how activation decisions are made:
$\text{IR} = |\{t : \text{trigger}(t) = \text{HOMEOSTATIC}\}| / |\{t : \text{action}(t) \neq \text{PASS}\}|$.
IR $\approx 1.0$ for HRRL (all active turns self-initiated) and IR $= 0.0$
for LangGraph (all active turns externally routed).

Taken together, these metrics provide a lightweight observability layer for
multi-agent discourse: all are computable in real time from the graph
structure, without human annotation or model-based scoring. This property is
increasingly valued as agentic systems scale and the need for automated
oversight grows~\parencite{wang2024survey}.

\subsubsection{Hypothesis.}

\begin{description}
    \item[H (Participation equity).] In a closed homeostatic activation
        loop where epistemic deficit is part of the agent's decision
        context, per-agent participation distributes significantly more
        uniformly than under exogenous LLM routing, under matched debate
        budgets.

        Formally: $H_0: \bar{\sigma}_{\text{HRRL}} \geq
        \bar{\sigma}_{\text{LG}}$; $H_1: \bar{\sigma}_{\text{HRRL}} <
        \bar{\sigma}_{\text{LG}}$, where $\sigma_s$ is the standard
        deviation of per-agent debate counts for seed $s$.

        Theoretical basis: symmetric homeostatic pressure---all agents
        accumulate deficit at the same basal rate, independent of
        faction---produces symmetric activation. An external router
        selecting by contextual relevance has no such symmetry guarantee;
        in an asymmetric discourse (attack vs.\ defense), it may
        systematically concentrate turns on the more contextually salient
        faction, replicating the dynamic Michels described.

        Falsification: no significant difference in per-agent participation
        variance across the five seeds.
\end{description}

Tested with a one-sided Welch's $t$-test on matched pairs ($n=5$ seeds).

% ============================================================
\section{Experimental Setup}
\label{sec:setup}

\begin{table}
\caption{Hyperparameters and their justification.}\label{tab:hyperparams}
\begin{tabular}{|l|l|l|p{5.5cm}|}
\hline
\textbf{Parameter} & \textbf{Symbol} & \textbf{Value} & \textbf{Justification} \\
\hline
Decay rate & $\lambda$ & 0.05 & Implies $\sim$1 debate per 4 heartbeats at equilibrium; comparable to decay constants in \parencite{laurencon2024ctcs} \\
Sigmoid steepness & $k$ & 10 & Gives 27\% activation at $\delta_0$ and 73\% at $\theta{+}0.1$: steep but smooth \\
Activation threshold & $\theta$ & 0.7 & Agents start at $\delta_0=0.5$; accumulate pressure for $\sim$4 heartbeats before first activation \\
Satiation rate & $\alpha$ & 2.0 & Design requirement: one good debate ($\Delta\varphi^*{\approx}0.15$) must reduce deficit below $\theta$; minimum $\alpha \geq 0.67$ \\
Stimulus gain & $\gamma$ & 0.1 & Relevance-scaled deficit increase per stimulus event \\
Baseline deficit & $\varepsilon$ & 0.1 & Homeostatic setpoint \\
Initial deficit & $\delta_0$ & 0.5 & Below $\theta$; ensures bootstrap phase \\
Drive exponent & $m$ & 2 & Keramati's quadratic case; superlinear pressure \\
Q-learning rate & $\eta$ & 0.01 & Conservative; avoids oscillation in weight updates \\
Discount factor & $\gamma_{RL}$ & 0.95 & Standard value; motivates shortest-path to setpoint \\
Max heartbeats (HRRL) & $T$ & 50 & LangGraph runs until matching debate count \\
Number of agents & $N$ & 4 & Two factions $\times$ two archetypes \\
LLM & --- & GPT-4o-mini (2024-07-18) & Pinned version for reproducibility \\
Temperature & --- & 0.7 & \\
Max tokens & --- & 300 & \\
Seeds & $s$ & $\{7, 17, 42, 123, 256\}$ & 5 seeds for Welch's $t$-test \\
\hline
\end{tabular}
\end{table}

\noindent The parameters $\theta$, $\lambda$, $\alpha$, and $k$ were set by
principled design requirements derived from the HRRL framework, not by
grid search. The $\Phi^*$ weights ($0.35$, $0.35$, $0.30$) reflect an
equal-importance prior; their optimization is a stated limitation.

Agent outputs are structured via Pydantic models enforcing a JSON schema:

\begin{lstlisting}
{"action": "DEBATE",
 "claim": "Las cifras contradicen ...",
 "target_node_id": "GOV-1_abc123",
 "attack_type": "rebuttal"}
\end{lstlisting}

Failed parses trigger a deficit penalty of $+0.1$ after two retry attempts.

The benchmark script runs both modes for each seed sequentially. A seed
factory derives deterministic prime seeds for each agent's RNG and LLM call
via SHA-256 hashing, ensuring reproducibility. We run five seeds
($s \in \{7, 17, 42, 123, 256\}$) and report mean $\pm$ standard deviation.
Statistical tests use one-sided Welch's $t$-test with Bonferroni correction.

\subsection{Research Question and Experimental Expectations}

The central research question is whether endogenous homeostatic regulation
produces more equitable participation dynamics than exogenous LLM-based
routing under identical task, model, seed, and debate-budget conditions---
and whether the Michels dynamic (contextual-relevance bias concentrating
influence) is observable in a neutral AI router.

Before running the benchmark, we expect the pattern stated in H: symmetric
homeostatic pressure should produce symmetric activation. All four agents
accumulate deficit at the same basal rate; faction-relevant stimuli
accelerate activation proportionally for the affected faction but not
structurally more for one faction than another. An external router has no
equivalent symmetry constraint: it selects by contextual relevance, and in
a political debate, the attacking faction's contributions are predictably
more contextually salient. We therefore expect HRRL to produce lower
per-agent participation variance ($\sigma_s$) than LangGraph. Additionally,
we expect HRRL deficits to oscillate around $\theta = 0.7$, exhibiting the
homeostatic cycle, whereas LangGraph deficits should grow monotonically,
reflecting the absence of endogenous satiation.

% ============================================================
\section{Results}
\label{sec:results}

\subsection{Benchmark Design}

We run the same simulation under HRRL and LangGraph using identical
configuration (GPT-4o-mini, $T = 50$ heartbeats for HRRL). Debate budgets
are matched: HRRL runs for $T$ heartbeats and produces $N$ debates
organically;
LangGraph then runs until it also reaches $N$ debate turns. We repeat the
experiment with five seeds ($s \in \{7, 17, 42, 123, 256\}$). This design
allows us to ask the data a constrained question: given the same debate
budget, which orchestration regime produces higher-quality argumentation?

\subsection{Structural Validity}

Before presenting outcome metrics, we verify that the two modes differ
structurally in the expected way. The Initiative Ratio (IR) is 1.0 for HRRL
and 0.0 for LangGraph across all five seeds, confirming that all active
turns under HRRL are self-initiated (trigger = HOMEOSTATIC) and all active
turns under LangGraph are externally chosen (trigger = ROUTER). This
manipulation check passes: the two modes are structurally distinct.

\subsection{Outcome Metrics}

\begin{table}
\caption{Comparative results (mean $\pm$ std across 5 seeds, matched debate
budget). HRRL: $T=50$ heartbeats, online Q-learning with TD(0). LangGraph:
runs until matching HRRL debate count.}\label{tab:results}
\begin{tabular}{|l|c|c|}
\hline
\textbf{Metric} & \textbf{HRRL} & \textbf{LangGraph} \\
\hline
Total debates              & 74.0 $\pm$ 9.6  & 74.0 $\pm$ 9.6 \\
Graph nodes                & 74.0 $\pm$ 9.6  & 74.0 $\pm$ 9.6 \\
Graph edges                & 72.8 $\pm$ 9.3  & 73.0 $\pm$ 9.6 \\
\hline
Per-agent $\sigma_s$ (\textbf{H}) & 1.53 $\pm$ 0.58 & 6.80 $\pm$ 1.14 \\
\hline
AAF defeat cycles          & 0 $\pm$ 0 & 0 $\pm$ 0 \\
AAF acceptance ratio       & 0.566 $\pm$ 0.021 & 0.711 $\pm$ 0.043 \\
PRR$_G$                    & 0.984 $\pm$ 0.004 & 0.986 $\pm$ 0.002 \\
Avg $\Delta\varphi^*$      & 0.149 $\pm$ 0.010 & 0.157 $\pm$ 0.011 \\
Avg reward (drive reduction) & 0.154 $\pm$ 0.015 & --- \\
\hline
IR (validity check)        & 1.0 & 0.0 \\
Stimulus-driven debates    & 63.6 $\pm$ 11.3 & --- \\
Avg stimulus utility       & 0.586 $\pm$ 0.031 & --- \\
\hline
\end{tabular}
\end{table}

Table~\ref{tab:results} presents the aggregate results. With both modes
producing the same number of debates by design (74.0 on average), five
patterns stand out.

\textbf{H confirmed: balanced participation as emergent property.} HRRL
produces a mean per-seed per-agent standard deviation $\bar{\sigma}_s =
1.53 \pm 0.58$, compared to $6.80 \pm 1.14$ for LangGraph ($t = -8.1$,
$df = 5.9$, $p < 0.001$, $d = -5.1$). The effect is large and consistent
across all five seeds. HRRL distributes debates nearly uniformly
($\sim$18.4 per agent), whereas LangGraph's router exhibits systematic
bias toward opposition agents (OPP: $\sim$25 vs.\ GOV: $\sim$12). This
$2.1\times$ imbalance reflects the router's tendency to favor the
attacking faction --- an instance of the Michels dynamic, where the
coordinator's contextual-relevance heuristic amplifies the informational
advantage of the more salient faction.

\textbf{Comparable argument quality.} HRRL achieves $\Delta\varphi^* =
0.149 \pm 0.010$, comparable to LangGraph's $0.157 \pm 0.011$. LangGraph
achieves marginally higher per-intervention quality; this difference is
not statistically significant at $\alpha = 0.05$ but is directionally
consistent across seeds ($d = -0.84$). The data suggests that endogenous
regulation achieves near-equivalent quality to centralized routing without
the overhead of an external selection function.

\textbf{More contested discourse under HRRL.} HRRL produces a
significantly lower AAF acceptance ratio ($\alpha = 0.566 \pm 0.021$)
than LangGraph ($\alpha = 0.711 \pm 0.043$). This means more arguments
are contested under HRRL. The pattern is consistent with balanced
participation: when all agents contribute equally, more claims face
counter-arguments. Under LangGraph, government agents are underrepresented,
so more opposition claims go uncontested, inflating $\alpha$. A lower
acceptance ratio signals a more thoroughly contested discourse, not
inferior quality.

\textbf{Stimulus-driven behavior.} 86\% of HRRL debates (63.6/74.0) are
triggered by specific discourse events. This confirms that the
event-driven architecture produces genuine stimulus-response behavior.

\textbf{Positive homeostatic reward.} The average reward (drive reduction)
per active heartbeat is $r = 0.154 \pm 0.015$, confirming that DEBATE
actions consistently reduce the epistemic drive, as predicted by the
HRRL equivalence theorem.

\subsection{Statistical Test}

\begin{table}
\caption{One-sided Welch's $t$-test ($n=5$ matched pairs) for H
(participation equity). Per-seed $\sigma_s$ = std.\ dev.\ of per-agent
debate counts within each seed.}\label{tab:stats}
\begin{tabular}{|l|c|c|c|c|}
\hline
\textbf{Hypothesis} & \textbf{$t$} & \textbf{$df$} & \textbf{$p$} &
\textbf{Cohen's $d$} \\
\hline
H: $\sigma_{\text{HRRL}} < \sigma_{\text{LG}}$ & $-8.1$ & 5.9 & $< 0.001$ & $-5.1$ \\
\hline
\end{tabular}
\end{table}

H is strongly confirmed ($t = -8.1$, $d = -5.1$): HRRL produces
significantly lower per-agent participation variance than LangGraph across
all five seeds (mean $\sigma_s$: $1.53$ vs.\ $6.80$). The effect size is
large, indicating that the symmetry of the homeostatic mechanism translates
reliably into balanced participation. This is consistent with the prediction
derived from Postulate~2 (drive symmetry) and the iron law framing: a neutral
external router concentrates turns on the contextually dominant faction;
endogenous regulation does not. These results are exploratory ($n=5$); a
post-hoc power analysis indicates $n \geq 12$ seeds would be needed to
detect $d = 0.5$ at 80\% power, though the observed $d = -5.1$ suggests the
effect is robust to seed variation.

\subsection{Q-Weight Convergence}

Across all 5 seeds and all 4 agents (20 agent--seed combinations), the
learned Q-weight for DEBATE\_STIMULUS exceeds that for DEBATE\_PROACTIVE
in every case (20/20). The mean final Q-values are: DEBATE\_STIMULUS
$= 0.247 \pm 0.076$, DEBATE\_PROACTIVE $= 0.186 \pm 0.044$, SEARCH
$= 0.034 \pm 0.039$, READ $= 0.017 \pm 0.019$. The Q-learner converges
to prioritize stimulus-driven debate over proactive exploration, consistent
with the prediction that a genuinely adaptive agent learns to respond to
faction-relevant threats rather than seeking diverse targets. The ordering
DEBATE\_STIMULUS $>$ DEBATE\_PROACTIVE $>$ SEARCH $>$ READ is stable
across seeds, providing evidence that the learning signal (drive reduction)
shapes action preferences in a consistent direction.

\subsection{Phase Analysis (HRRL)}

The HRRL simulation exhibits three phases, consistent with the theoretical
prediction:

\textbf{Phase 1: Bootstrap.} Heartbeats 1--3: agents initiate proactive
debates to seed the argument graph. Deficits grow from $\delta_0 = 0.5$
toward $\theta = 0.7$ via basal decay.

\textbf{Phase 2: Stimulus Onset.} Around heartbeat 3--5: as the graph
grows, \texttt{NewArgumentEvent}s begin flowing. Agents stimulate: deficit
increases proportionally to stimulus relevance
(Eq.~\ref{eq:stimulate}). This accelerates the crossing of $\theta$ for
agents whose faction is under attack.

\textbf{Phase 3: Homeostatic Oscillation.} Heartbeats 5--50: deficits
oscillate around $\theta$ (final means: 0.50--0.66). Acting reduces the
deficit via satiation, which reduces activation probability, which allows
the deficit to grow again through decay and stimulation. This is the
homeostatic cycle emerging without external intervention. Concurrently, the
Q-learner accumulates positive reward ($\bar{r} = 0.154$) from each
successful drive reduction, reinforcing the DEBATE\_STIMULUS action whose
weight grows relative to alternatives.

Under LangGraph, deficit dynamics diverge: deficits grow monotonically
because agents are externally routed and satiation is insufficient to
counteract continuous decay. Final LangGraph deficits exceed 2.1--3.6,
far above $\theta$, indicating the absence of self-regulation.

% ============================================================
\section{Discussion}
\label{sec:discussion}

\subsection{Interpreting the Results}

The results paint a nuanced picture. HRRL achieves equivalent argument
quality to LangGraph while producing structurally different---and in some
respects superior---participation dynamics. We address each finding in turn.

\textbf{Comparable quality with genuine learning.} HRRL achieves
$\Delta\varphi^* = 0.149 \pm 0.010$, comparable to LangGraph's
$0.157 \pm 0.011$ ($d = -0.84$). The Q-learner adapts action selection
online: by the end of the simulation, DEBATE\_STIMULUS consistently
receives the highest Q-value ($0.247 \pm 0.076$), confirming that the
agent learns to prioritize stimulus-driven engagement over proactive
exploration. The average homeostatic reward per active heartbeat
($r = 0.154$) is positive, confirming that DEBATE actions consistently
reduce drive, as predicted by the equivalence theorem. Convergence within
50 heartbeats is limited; the learned weights should be interpreted as an
adaptive prior.

\textbf{Balanced participation as an emergent property (H confirmed).}
HRRL produces mean per-seed $\bar{\sigma}_s = 1.53$, compared to $6.80$
for LangGraph ($p < 0.001$, $d = -5.1$). HRRL distributes debates nearly
uniformly across agents, whereas LangGraph concentrates activity on
opposition agents (OPP receiving $2.3\times$ more turns than GOV). This
imbalance is consistent with Michels' iron law applied to AI orchestration:
the external router, acting as neutral coordinator, selects agents whose
contributions appear most contextually relevant---and in a political debate,
the attacking faction's arguments consistently surface more salient context.
This is not a flaw in the router's design; it is the structural consequence
of centralized selection in an asymmetric discourse. Endogenous regulation
counters this tendency through symmetric homeostatic pressure: all agents
accumulate deficit at the same basal rate, independent of faction, so no
single faction accrues structural informational advantage in the activation
decision. Balanced participation is not engineered---it is an emergent
consequence of the symmetry of the regulatory mechanism.

\textbf{No defeat cycles in either mode.} The absence of AAF defeat cycles
in both modes reflects a structural feature of the debate rather than a
limitation of the orchestration mechanism. In a two-faction political debate,
the discourse follows an asymmetric attack pattern (opposition attacks
government claims; government defends). Symmetric refutation cycles would
require agents to explicitly counter-attack claims that already attacked
them, which the current prompting strategy does not encourage.

\textbf{Homeostatic oscillation confirmed.} HRRL deficits oscillate around
$\theta = 0.7$ (final means: 0.50--0.66), exhibiting the predicted
three-phase behavior. LangGraph deficits grow monotonically (final means:
2.1--3.6), confirming the absence of self-regulation.

From a broader perspective, the contrast is not simply between two
implementation choices but between two theories of coordination. One holds
that coordination is best imposed by an external controller with full
information. The other holds that coordination can emerge from local
regulation. The present results suggest that endogenous regulation achieves
equivalent quality to centralized routing while producing emergent properties
---balanced participation, homeostatic oscillation, stimulus-responsive
activation---that the centralized approach cannot produce by design.

A reasonable objection is that the deficit mechanism reduces to a dynamic
priority queue with self-adjusting weights. The Q-learner makes the
distinction sharper: a priority queue cannot learn; the deficit-driven
agent does. The Q-weights adapt online based on which actions most
effectively reduce drive, producing behavior that changes over the
course of the simulation. Priority queues rank; Q-learners adapt.

A related objection is that the mechanism resembles contextual bandits.
The distinction is structural: unlike bandits, the agent's state (deficit)
evolves endogenously through the drive function (Eq.~\ref{eq:drive}), and
the reward depends on state transitions (Eq.~\ref{eq:reward}), not just
action outcomes. The temporal discount factor $\gamma$ and the TD update
make this a sequential decision problem, not a stateless one.

A candid observation: we describe HRRL as ``endogenous'' regulation, but
the regulatory mechanism was designed exogenously by the researchers. The
deficit, threshold, and satiation parameters are designer choices---a
``mapmaker,'' in the terminology of~\textcite{goo2025abstraction}. What is
endogenous is the \emph{dynamics}: once the mechanism is in place, the
oscillatory equilibrium emerges from agent-environment interaction without
further external intervention. This distinction keeps the claim honest.

This rupture can be stated precisely. As discussed in
Section~\ref{sec:related}, both the heartbeat
paradigm~\parencite{openclaw2026heartbeat} and LangGraph-style routing
externalize the activation decision: the agent does not control its own
pulse. LangClaw internalizes it. The heartbeat signal still fires
uniformly, but the agent's \emph{propensity to act} is governed by its
deficit. The result, confirmed empirically, is that participation frequency
becomes an emergent variable rather than a design constant.

A complementary interpretation frames HRRL as an adaptive inference-time
compute allocator. Recent scaling laws for test-time
compute~\parencite{snell2024scaling} have established that performance
improves predictably when more tokens are invested at inference. In a
multi-agent setting, this raises a distributional question: given a fixed
compute budget, how should LLM calls be allocated across agents and time?
Centralized routing answers via a global selection function (one additional
LLM call per heartbeat). HRRL answers adaptively: agents consume compute only
when their stimulus-evaluated internal state warrants it, with no routing
overhead. Our results show that HRRL achieves equivalent quality to
centralized routing while distributing compute more uniformly across agents.
Under tighter compute budgets or larger agent populations, this
self-balancing property may become a decisive advantage.

\subsection{From AopifyJS to LangClaw}

LangClaw extends the agent-based architecture of
AopifyJS~\parencite{gerpe2023aopifyjs} along the three open directions
identified in that work: human-like motivational states (now implemented via
the epistemic drive), formal argumentation semantics (now operationalized
through the argument graph with undercut and rebuttal relations), and
reinforcement learning integration (now provided by the drive function,
reward-as-drive-reduction, and online Q-learning with TD(0) updates). The
event-driven communication pattern---asynchronous queues, self-contained
state, structured interaction history---is preserved.

\subsection{Limitations}

We identify eight limitations:

\begin{enumerate}
    \item \textbf{Statistical power}: A post-hoc power analysis indicates
          $n \geq 12$ seeds would be needed to detect $d = 0.5$ at 80\%
          power; our $n=5$ is exploratory. Results should be interpreted as
          preliminary evidence, not confirmatory findings.
    \item \textbf{Q-learning convergence}: With 50 heartbeats and
          $\eta = 0.01$, the Q-learner does not reach convergence. The
          learned weights should be interpreted as an adaptive prior that
          improves upon the warm-start heuristic, not as a converged policy.
    \item \textbf{Defeat cycle absence}: Neither mode produced AAF defeat
          cycles. The prompting strategy does not explicitly encourage
          symmetric counter-attacks; future work should address this.
    \item \textbf{PRR text inapplicability}: The text-based PRR metric
          yields zero for both modes because agents argue against positions
          rather than naming peers by ID. The graph-structural variant
          (PRR$_G$) was reported instead.
    \item \textbf{$\Phi^*$ weights}: The weights ($0.35, 0.35, 0.30$)
          reflect an equal-importance prior and were not optimized.
    \item \textbf{LLM dependence}: Results are produced with GPT-4o-mini
          (version 2024-07-18) and will vary with different models,
          temperatures, and prompt formulations.
    \item \textbf{Scale}: Four agents and 50 heartbeats is small. Behavior
          with more agents or longer horizons is untested.
    \item \textbf{Safety and externalities}: Self-regulating agents operate
          without external oversight during execution. The deficit mechanism
          could be gamed: agents that produce structurally well-placed but
          semantically vacuous arguments would still receive satiation.
\end{enumerate}

\subsection{Further Research}

Several directions follow from this work. First, prompting strategies that
encourage symmetric counter-attacks would make defeat cycles testable.
Second, hyperparameter sensitivity studies for $\theta$, $\alpha$,
$\lambda$, and $\gamma$ would characterize robustness. Third, scaling to
larger agent populations ($N > 4$) and longer horizons would test whether
balanced participation holds at scale. Fourth, replacing the heuristic
$\Phi^*$ proxy with a learned quality function would improve the fidelity of
the homeostatic signal. Fifth, scaling the Q-learner across episodes
(transfer of learned weights across debate topics) and learning the drive
exponent $m$ online are natural extensions. Finally, evaluating HRRL under
constrained inference-time compute budgets, where the adaptive allocation of
LLM calls may become a decisive advantage over uniform scheduling, is a
particularly promising direction.

% ============================================================
\section{Conclusion}
\label{sec:conclusion}

This paper examined a concrete question: whether context collapse in
multi-agent LLM debate is better approached through external scheduling or
through endogenous regulation. We introduced LangClaw as an experimental
framework for studying that question under controlled conditions, comparing a
homeostatic mechanism against a LangGraph-style exogenous router with matched
debate budgets, while reducing role-related confounds through symmetric agent
design.

The relevance of this question extends beyond the present prototype. If
coherence in multi-agent interaction depends mainly on external control, then
future systems will likely continue to evolve toward more sophisticated
routing and supervision layers. If, by contrast, part of the problem can be
addressed through endogenous regulation, then a different opportunity
emerges: multi-agent systems may be designed less as scripted workflows and
more as adaptive cognitive ecologies. This distinction matters for the
development of socially relevant applications in domains such as policy
analysis, scientific assistance, education, and other forms of persistent AI
coordination.

Our results show that the answer is not ``either/or'' but depends on what
properties of coordination matter. Endogenous regulation achieves equivalent
argument quality to centralized routing while producing emergent structural
advantages---balanced participation, homeostatic oscillation, and
stimulus-responsive activation---that centralized approaches cannot produce
by design.

The contributions of this paper are:

\begin{enumerate}
    \item A formal specification of a bio-inspired, event-driven
          homeostatic activation loop for multi-agent LLM systems ---
          an extension of the OpenClaw paradigm with epistemic deficit as
          an internal context signal --- grounded in allostatic regulation
          and eudaimonic wellbeing, and integrating online Q-learning with
          homeostatic reward.
    \item A multi-criteria stimulus evaluator that enables agents to assess
          incoming discourse events for faction relevance, target centrality,
          strategic memory, novelty, and argumentative pressure, with
          deficit-modulated execution probability.
    \item A network-theoretic argument quality proxy $\Delta\varphi^*$ and a
          comparative benchmark framework with matched debate budgets, shared
          memory infrastructure, and symmetric agent roles, enabling
          controlled comparison of endogenous and exogenous orchestration.
    \item Experimental evidence that endogenous homeostatic regulation
          produces significantly more balanced agent participation than
          centralized LLM routing (mean per-seed $\sigma$: $1.53$ vs.\
          $6.80$, $t \approx -8.1$, $p < 0.001$, $d \approx -5.1$),
          consistent with the hypothesis that symmetric homeostatic pressure
          counters the concentrating dynamic Michels described in human
          deliberative organizations.
\end{enumerate}

% ============================================================
\section*{Reproducibility}

The complete codebase is available at:
\url{https://github.com/pgerpe/langclaw_experiment}

\begin{verbatim}
pip install -r requirements.txt
python main.py --model gpt-4o-mini --iterations 50
python benchmark.py --modes hrrl langgraph --seeds 7 17 42 123 256 --iterations 50
\end{verbatim}

% ============================================================
\begin{credits}
\subsubsection{\discintname}
The authors have no competing interests to declare that are relevant to the
content of this article.
\end{credits}

% ============================================================
\printbibliography

\end{document}
